---
title: "R Hack – Getting started with maps in R using GeoJSON and library sf"
subtitle: "A simple workflow to load and plot maps in R"
description: "R-Hacks"
author: "Lucio Colonna"
date: "05 December 2025"
images: "images/image-1.png"
series: R-Hacks
categories:
  - GeoJSON
  - ggplot
  - maps
layout: single
toc-location: left
toc: true
code-overflow: wrap
execute: 
  warning: false
  message: false
---

*This hack is based on my Cyclistic analysis on Kaggle (see Chapter **8.12.4**):*  
🔗 <https://www.kaggle.com/code/lcolon/cyclistic-2023-google-da-capstone-project-r>

Working with maps in R is easier than it seems.  
If you want to visualize points, regions, or spatial patterns, the combination of **GeoJSON files** and the **`sf`** package gives you a simple and modern workflow.

In this hack, we’ll load a real city map (Chicago) directly from a URL, plot it with `geom_sf()`, and overlay simple point data.

---

## Step 0 – What is a GeoJSON file?

A **GeoJSON** file is a **JSON text file** that stores **geographic shapes** (points, lines, polygons), together with their **attributes** (names, boundaries, codes, etc.).

It is one of the most common formats for open geographic data because it is **lightweight, human-readable, and web-friendly**.

You can often find GeoJSON files on:

- city open data portals (City of Chicago, NYC Open Data, etc.)  
- national and international platforms (e.g., data.gov)  
- GitHub repositories  
- many public mapping resources

---

## Step 1 – Load the required packages

```{r}
library(sf)       # for spatial data (simple features)
library(ggplot2)  # for plotting
```

---

## Step 2 – Read a GeoJSON map with `sf::st_read()`

In this hack, we use a **GeoJSON of Chicago community areas**, provided by the [City of Chicago Open Data Portal](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Neighborhoods/bbvz-uum9).

*⚠️ Before using any online map or dataset, always remember to check the terms of use of the source.*

You can read a GeoJSON directly from a URL, without downloading anything manually:

```{r}
map_of_chicago <- sf::st_read(
"https://data.cityofchicago.org/resource/y6yq-dbs2.geojson",
quiet = TRUE
)

head(map_of_chicago)
```

---

What you get:

- `map_of_chicago` is an sf object -essentially a data frame with a special geometry column where each row is a polygon that represents one community area

---

## Step 3 – Plot a basic map with `geom_sf()`

Once the GeoJSON file has been loaded into an sf object, you can visualize it with `geom_sf()`.

`geom_sf()` is the ggplot2 layer designed specifically for spatial data: it knows how to read the geometry column of an sf object and automatically draw points, lines, or polygons depending on what the data contains.
In our case, it will draw the polygons of Chicago’s community areas.

```{r}
ggplot() +
geom_sf(data = map_of_chicago, fill = "#f1f0f0", color = "grey50") +
labs(
title = "Chicago community areas",
subtitle = "GeoJSON map loaded with sf::st_read()"
) +
theme_void()
```

---

# Step 4 – Add your own data on top of the map

If you have a dataset with latitude/longitude, you can overlay it on the map.

For example, let’s create a small dummy dataset of bike sharing station locations:

```{r}
stations <- data.frame(
  start_lng = c(
    -87.6278, -87.6560, -87.6705, -87.6045, -87.5950,
    -87.6850, -87.6515, -87.7080, -87.6405, -87.5600
  ),
  start_lat = c(
    41.8925, 41.9000, 41.9120, 41.8560, 41.7920,
    41.9220, 41.9250, 41.8810, 41.9410, 41.7600
  )
)

stations

```

Now we can plot the base map plus the stations:

```{r}
ggplot() +
geom_sf(data = map_of_chicago, fill = "#f1f0f0", color = "grey70") +
geom_point(
data = stations,
aes(x = start_lng, y = start_lat),
color = "red",
size = 3
) +
labs(
title = "Bike stations on top of a GeoJSON map of Chicago",
subtitle = "Base map: Chicago community areas (GeoJSON + sf)"
) +
theme_void()
```

```{r}
#| eval: false
#| echo: false
library(ggplot2)
ggsave("images/image-1.png", 
       bg = "white",
       width = 6, height = 4)
```


---

## To Recap

- GeoJSON = JSON file with geometry + attributes

- Load it directly from a URL with `sf::st_read()`

- Plot it with `geom_sf()`

- Add your own points using regular ggplot layers

- This workflow is simple, beginner-friendly, and requires no GIS tools

<br>

**Thanks for reading!**

::: callout-tip
If you want to stay up to date with the latest events from the **Rome R Users Group**, click here:

👉 <https://www.meetup.com/rome-r-users-group/>

And if you are curious, the full Kaggle notebook used for this tip is available here:

🔗 <https://www.kaggle.com/code/lcolon/cyclistic-2023-google-da-capstone-project-r>
:::
